Scattering from compact objects: \\ Regge poles and the Complex Angular Momentum method

### 2019-12-13 rainbowCAM.tex
First complete draft for a read-through.

\affiliation{Consortium for Fundamental Physics,  School of Mathematics and Statistics,
University of Sheffield, Hicks Building, Hounsfield Road, Sheffield S3 7RH, United Kingdom \looseness=-1}
%
\author{Tom Stratton}\email{tstratton1@sheffield.ac.uk}
\affiliation{Consortium for Fundamental Physics,  School of Mathematics and Statistics,
University of Sheffield, Hicks Building, Hounsfield Road, Sheffield S3 7RH, United Kingdom \looseness=-1}
%
\author{Sam R. Dolan}\email{s.dolan@sheffield.ac.uk}
\affiliation{Consortium for Fundamental Physics,  School of Mathematics and Statistics,
University of Sheffield, Hicks Building, Hounsfield Road, Sheffield S3 7RH, United Kingdom \looseness=-1}
%
\begin{abstract}
We calculate the Regge poles of the scattering matrix for a compact gravitating body, showing that the spectrum has three branches, which we classify as surface waves, broad resonances and narrow resonances. We examine the spectrum for neutron-star-like body, and compare with the spectrum for an ultra-compact body.  for the massless scalar field and the odd-parity gravitational-wave sector.
\end{abstract}

@@

CAM theory, originally developed for high-energy physics applications, has been successfully applied to black hole scattering scenarios. In 1994, Andersson and Thylwe \cite{Andersson:1994rk} described the scattering of scalar waves by a Schwarzschild BH by utilising CAM techniques, and they showed that, in this case, Regge poles are associated with ``surface waves'' localised near the unstable light ring at $r=3M$ (later it was shown that the corresponding quasinormal modes are Breit-Wigner type resonances generated by these surface waves \cite{Decanini:2002ha}). Recently, Folacci and Ould Elhadj have shown that CAM methods can be used for precise numerical calculations of scattering cross sections of scalar fields, electromagnetic fields \cite{Folacci:2019cmc} and gravitational waves \cite{Folacci:2019vtt} on the Schwarzschild spacetime. Moreover, they have obtained an approximate  formula (see Sec.~IV in Ref.~\cite{Folacci:2019vtt}) that accurately describes both the `orbiting' and `glory' oscillations in black hole scattering cross sections in the short-wavelength regime. This approximation arises from applying WKB theory to characterise the Regge pole with the smallest imaginary part, and its residue \sam{Mohamed, please correct this if wrong.}. An accurate approximation for the black hole absorption cross section has also been derived via an analysis of Regge poles \cite{Decanini:2011xi, Decanini:2011xw}.

@@

A widely-studied model is that of a polytropic star, with an EoS $p(\rho) = \kappa \rho^{1+1/\hat{n}}$, where $\hat{n}$ is the polytropic index (see e.g.~\cite{Stratton:2019deq}). Here we shall consider a special case: an incompressible perfect fluid ball of uniform density described by Schwarzschild's interior solution for an incompressible fluid \cite{Shapiro1983}, with
\begin{subequations}
\begin{align}
\rho &= \frac{M}{\frac{4}{3} \pi R^3} , \\
p &= \rho \frac{\beta(R) - \beta(r)}{ \beta(r) - 3 \beta(R)} , \\
\beta(x) &= \sqrt{3 - 8 \pi \rho x^2}
\end{align}
\end{subequations}
and metric functions
%
\begin{subequations}\label{Interior_Solution}
\begin{align}\label{Interior_Solution_f}
 f(r) &= \frac{1}{4}\left(1-\frac{2 M r^2}{R^3}\right)+\frac{9}{4}\left(1-\frac{2M}{R}\right) \nn \\
       & \quad -\frac{3}{2} \sqrt{\left(1-\frac{2M}{R}\right)\left(1-\frac{2 M r^2}{R^3}\right)}, \\
 h(r) &= 1-\frac{2 M r^2}{R^3} . \label{Interior_Solution_h}
\end{align}
\end{subequations}
The constant-density model can be thought of as representing the $\hat{n} \rightarrow 0$ limit of the family of polytropes.
The radial function $h(r)$ is $C^0$ at the surface of the star $r=R$ (i.e.~continuous but not differentiable), and the radial function $f(r)$ is $C^1$ there (i.e.~once-differentiable).
%In the polytropic cases $n > 0$, the functions are more regular \sam{can we make a precise statement?}\tom{I will write something here} at the surface, but not $C^\infty$ (i.e.~not smooth). As we shall see, the breakdown of smoothness leads to consequences for the Regge pole spectrum.
For a general polytrope with $n>0$ and $m$ the smallest integer such that $m \geq n$, $h(r)$ is $C^m$ and $f(r)$ is $C^{m+1}$ at the surface.
%\tom{This can be seen from equations relating $(m,\rho,p,f,h)$ and by noting $\rho \sim a(R-r)^n$ at the surface \cite{Lindblom:1983ps} but we don't mention these.}
As we shall see, the breakdown of smoothness leads to consequences for the Regge pole spectrum.

@@

%It is important to note that the poles of  $S_{\lambda -1/2} (\omega)$ in the complex $\lambda$ plan (i.e., the Regge poles) are defined  as  the zeros of the coefficient  $A^{(-)}_{\lambda-1/2} (\omega)$ [see Eq.~(\ref{Matrix_S_CAM})], i.e., the values $\lambda_n(\omega)$ such that
%\begin{equation}\label{PR_def_Am}
%A^{(-)}_{\lambda_n(\omega)-1/2} (\omega)=0,
%\end{equation}
%with $n=1,2,3,\ldots$
%The residue of the matrix $S_{\lambda-1/2}(\omega)$ at the pole $\lambda=\lambda_n(\omega)$ is defined by [see Eq.~(\ref{Matrix_S_CAM})]
%\begin{equation}\label{residues_RP}
%r_n(\omega)=e^{i\pi [\lambda_n(\omega)+1/2]} \left[ \frac{A_{\lambda -1/2}^{(+)}(\omega)}{\frac{d}{d \lambda}A_{\lambda -1/2}^{(-)}(\omega)}\right]_{\lambda=\lambda_n(\omega)}.
%\end{equation}
%These residues play a central role in the complex angular momentum paradigm.
%
\moh{Remark for Tom and Sam : In the next paragraph, I modified it to include the Regge poles of the third quadrant by symmetry to the origin $O$ , in the others words by changing $\ell \rightarrow -\ell-1$ } \\

@@

\begin{figure*}[htp]
\centering
 \includegraphics[scale=0.50]{Scattering_Cross_Section_R_2-dot-26_2Mw_3}
  \vspace*{-0.3cm}
\caption{\label{S_0_R_2-dot-26_2Mw_3_Exact_vs_CAM} The scalar cross section of a very compact bodies for $2M\omega=3$ and $R=2.26M$ and its Regge pole approximation.  The plots show the effect of including successively more Regge poles (plots 1--3). In the final plot, The sum over Regge poles gives a cross section which agrees well with the (regularized) partial-wave sum for intermediate and large values of the scattering
angle.}
\end{figure*}

\begin{figure*}[htp]
\centering
 \includegraphics[scale=0.50]{Scattering_Cross_Section_R_2-dot-26_2Mw_6}
  \vspace*{-0.3cm}
\caption{\label{S_0_R_2-dot-26_2Mw_6_Exact_vs_CAM} The scalar cross section of a very compact bodies for $2M\omega=6$ and $R=2.26M$ and its Regge pole approximation. The plots show the effect of including successively more Regge poles (plots 1--3). In the final plot, The sum over Regge poles gives a cross section which agrees well with the (regularized) partial-wave sum for intermediate and large values of the scattering
angle.}
\end{figure*}

@@

Figures \ref{Rainbow_Cross_Section_R_6_2Mw_16} and \ref{Diff_Contribution_Cross_Section_R_2-dot-26_2Mw_6} show the relative magnitudes of the contributions from the different branches of Regge poles. For $R/M = 6$ (Fig.~ \ref{Rainbow_Cross_Section_R_6_2Mw_16}) the surface-wave and broad-resonance amplitudes are similar in magnitude, and similar in magnitude to their sum. There is a difference in phase between these amplitudes which generates the peaks and troughs of the rainbow-scattering pattern.

@@

In this work, we have: (1) calculated the full spectrum of Regge poles for a compact body spacetime for the first time; (2) applied CAM theory to obtain an expression for the scattering amplitude as a residue series accompanied by a background integral, Eqs.~(\ref{CAM_Scalar_Scattering_amp_tot})--(\ref{CAM_Scalar_Scattering_amp_decomp_RP}); and (3) utilised the CAM formulae to calculated scattering cross sections numerically, demonstrating precise agreement with the numerical results of the partial-wave expansion first computed in Ref.~\cite{Dolan:2017rtj}.

@@

By applying a WKB method at lowest order, we have obtained a better understanding of the spectrum of the broad resonances, finding that the magnitude of the imaginary part is linked to the magnitude of the discontinuity in the effective potential at the surface of the body (see Eq.~(\ref{eq:alpha}), Eq.~(\ref{lambda_Approx}) and Fig.~\ref{fig:Veff}). Taking the WKB method to higher orders would improve the accuracy of approximation. Further work is need to obtain asymptotic approximations for the surface-wave branch, and the narrow-resonance branch.

We can be confident that there are no additional branches Regge poles in the first quadrant, for two reasons. First, we have scanned across the complex-$\lambda$ domains shown in Figs.~\ref{RP_R6_approx_2Mw_3_6_s_1} and \ref{RP_R226_approx_2Mw_3_6_s_1}, as detailed in Sec.~\ref{subsec:method}. Second, in our calculation of $d\sigma/d\Omega$, we find a precise numerical agreement between the CAM calculation and the partial-wave calculation; this would not be the case if poles in this domain had been missed.

The CAM toolkit was applied here to gain a complementary practical method for calculating scattering amplitudes. However, in its current form, the method is not particularly efficient, for two reasons; first, it requires the computation of a background integral in addition to residue sum; and second, the poles on the `broad resonance' branch have an approximately uniform spacing $\Delta \lambda \sim 2$, and thus a large number of poles must be found. In the black-hole case, by contrast, there is only a single branch of surface-wave-type modes (see Fig.~\ref{RP_R6_approx_2Mw_3_6_s_1}), and in the high-frequency limit, a single pole captures both the orbiting and glory phenomena \cite{Folacci:2019cmc, Folacci:2019vtt}. In the case of Mie scattering, a powerful extension of the CAM approach is to define the so-called Regge-Debye poles \cite{Nussenzveig:2006}, thereby eliminating the surface-wave branch entirely. This is a possible direction for future investigation.

%In this work, we have  $R=2.26M$, which is above the Buchdahl bound but nevertheless below ... blah blah, work in Tom's text.

%Remarkably, in the 1970s, experimental data on rainbows in nucleon-nucleon scattering experiments was used to constrain models of the nuclear potential.

%Finally a paragraph on the connections to Mie scattering by a transparent sphere.

\acknowledgments
M.O.E.H. wishes to thank Antoine Folacci for various discussions concerning this work and thanks Sam R. Dolan for his kind
invitation to the University of Sheffield. T.S.~acknowledges financial support from EPSRC. S.R.D.~acknowledges financial support from the European Union's Horizon 2020 research and innovation programme under the H2020-MSCA-RISE-2017 Grant No.~FunFiCO-777740, and from the Science and Technology Facilities Council (STFC) under Grant No.~ST/P000800/1.

>>>>

\affiliation{Consortium for Fundamental Physics,  School of Mathematics and Statistics,
University of Sheffield, Hicks Building, Hounsfield Road, Sheffield S3 7RH, United Kingdom \looseness=-1}
%
\author{Tom Stratton}\email{tstratton1@sheffield.ac.uk}
\affiliation{Consortium for Fundamental Physics,  School of Mathematics and Statistics,
University of Sheffield, Hicks Building, Hounsfield Road, Sheffield S3 7RH, United Kingdom \looseness=-1}
%
\author{Sam R. Dolan}\email{s.dolan@sheffield.ac.uk}
\affiliation{Consortium for Fundamental Physics,  School of Mathematics and Statistics,
University of Sheffield, Hicks Building, Hounsfield Road, Sheffield S3 7RH, United Kingdom \looseness=-1}
%
\begin{abstract}
We calculate the Regge poles of the scattering matrix for a gravitating compact body, for scalar fields and for gravitational waves in the axial sector. For a neutron-star-like body, the spectrum exhibits two distinct branches of poles, labelled surface waves and broad resonances; for ultra-compact objects, the spectrum also includes a finite number of narrow resonances. We show, via a WKB analysis, that the discontinuity of the effective potential at the body's surface determines the imaginary component of the broad-resonance poles.

Next, we examine the role of Regge poles in the time-independent scattering of monochromatic planar waves. We apply Complex Angular Momentum techniques to re-sum the partial wave series for the scattering amplitude, expressing it as a residue series evaluated at poles in the first quadrant, accompanied by a background integral. We compute the scattering cross section at several frequencies, and show precise agreement with the partial-wave calculations. Finally, we show that compact bodies naturally give rise to orbiting, glory, and rainbow-scattering interference effects.
\end{abstract}

@@

%CAM theory, originally developed for high-energy physics applications, has been successfully applied to black hole scattering scenarios. In 1994, Andersson and Thylwe \cite{Andersson:1994rk} described the scattering of scalar waves by a Schwarzschild BH by utilising CAM techniques, and they showed that, in this case, Regge poles are associated with ``surface waves'' localised near the unstable light ring at $r=3M$ (later it was shown that the corresponding quasinormal modes are Breit-Wigner type resonances generated by these surface waves \cite{Decanini:2002ha}). Recently, Folacci and Ould Elhadj have shown that CAM methods can be used for precise numerical calculations of scattering cross sections of scalar fields, electromagnetic fields \cite{Folacci:2019cmc} and gravitational waves \cite{Folacci:2019vtt} on the Schwarzschild spacetime. Moreover, they have obtained an approximate  formula (see Sec.~IV in Ref.~\cite{Folacci:2019vtt}) that accurately describes both the `orbiting' and `glory' oscillations in black hole scattering cross sections in the short-wavelength regime. This approximation arises from applying WKB theory to characterise the Regge pole with the smallest imaginary part, and its residue \sam{Mohamed, please correct this if wrong.}. An accurate approximation for the black hole absorption cross section has also been derived via an analysis of Regge poles \cite{Decanini:2011xi, Decanini:2011xw}.

The CAM theory was originally developed to deal theoretically with the propagation and diffraction of radio waves around the Earth, by Watson~\cite{Watson18} (see also the work of Sommerfeld~\cite{Sommerfeld49}). It has since been extensively used in several domains of physics involving resonant scattering theory (see, \textit{e.g.}, Refs.~\cite{deAlfaro:1965zz,Newton:1982qc,Watson18,Sommerfeld49,Nus92,Grandy2000,Uberall1992,AkiRichards2002,AkiRichards2002,Gribov69,
Collins77,BaronePredazzi2002,DonnachieETAL2005} as well as references therein for various applications in quantum mechanics, nuclear physics, electromagnetism, optics, seismology and high energy physics). Since the pioneering work in 1994 by Andersson and Thylwe~\cite{Andersson:1994rk,Andersson:1994rm} the CAM theory has been successfully applied to black hole scattering scenarios. They described the scattering of scalar waves by a Schwarzschild black hole using CAM techniques, and showed that, in this case, the Regge poles are associated with ``surface waves'' localised near the unstable light ring at $ r = 3M $ (\textit{i.e.}, the so-called photon sphere). Later, Decanini, Folacci and Jensen have shown that the complex frequencies of weakly damped QNMs are Breit-Wigner resonances generated by surface waves previously mentioned and, using the concept of Regge trajectories, they were able to construct semiclassically the spectrum of the QNM complex frequencies~\cite{Decanini:2002ha}. Recently, Folacci and Ould~El~Hadj have shown that CAM machinery can be used for precise numerical calculations of scattering cross sections of scalar fields, electromagnetic fields~\cite{Folacci:2019cmc} and gravitational waves~\cite{Folacci:2019vtt} on Schwarzschild space-time. Moreover, using the third WKB approximation to obtain the asymptotic expressions for the lowest Regge poles~\cite{Decanini:2009mu} and associated residues, they have able to provide an analytical approximation describing accurately both ``glory'' and large part of the ``orbiting'' oscillations in black hole scattering cross sections in the short-wavelength regime. An accurate approximation for the black hole absorption cross section has also been derived via an analysis of Regge poles~\cite{Decanini:2011xi,Decanini:2011xw}. In addition, an alternative description of gravitational radiation from black holes based on CAM theory was developed in Ref.~\cite{Folacci:2018sef}.

@@

A widely-studied model is that of a polytropic star, with an EoS $p(\rho) = \kappa \rho^{1+1/\hat{n}}$, where $\hat{n}$ is the polytropic index (see e.g.~\cite{Stratton:2019deq}). Here we shall consider a special case: an incompressible perfect fluid ball of uniform density described by Schwarzschild's interior solution for an incompressible fluid \cite{Shapiro1983}, with
\begin{subequations}
\begin{align}
\rho &= \frac{M}{\frac{4}{3} \pi R^3} , \\
p &= \rho \frac{\beta(R) - \beta(r)}{ \beta(r) - 3 \beta(R)} , \\
\beta(x) &= \sqrt{3 - 8 \pi \rho x^2}
\end{align}
\end{subequations}
and metric functions
%
\begin{subequations}\label{Interior_Solution}
\begin{align}\label{Interior_Solution_f}
 f(r) &= \frac{1}{4}\left(1-\frac{2 M r^2}{R^3}\right)+\frac{9}{4}\left(1-\frac{2M}{R}\right) \nn \\
       & \quad -\frac{3}{2} \sqrt{\left(1-\frac{2M}{R}\right)\left(1-\frac{2 M r^2}{R^3}\right)}, \\
 h(r) &= 1-\frac{2 M r^2}{R^3} . \label{Interior_Solution_h}
\end{align}
\end{subequations}
The constant-density model can be thought of as representing the $\hat{n} \rightarrow 0$ limit of the family of polytropes.
The radial function $h(r)$ is $C^0$ at the surface of the star $r=R$ (i.e.~continuous but not differentiable), and the radial function $f(r)$ is $C^1$ there (i.e.~once-differentiable).
For a general polytrope with $n>0$ and $m$ the smallest integer such that $m \geq n$, $h(r)$ is $C^m$ and $f(r)$ is $C^{m+1}$ at the surface. 
As we shall see, the breakdown of smoothness leads to consequences for the Regge pole spectrum.

@@

%It is important to note that the poles of  $S_{\lambda -1/2} (\omega)$ in the complex $\lambda$ plan (i.e., the Regge poles) are defined  as  the zeros of the coefficient  $A^{(-)}_{\lambda-1/2} (\omega)$ [see Eq.~(\ref{Matrix_S_CAM})], i.e., the values $\lambda_n(\omega)$ such that
%\begin{equation}\label{PR_def_Am}
%A^{(-)}_{\lambda_n(\omega)-1/2} (\omega)=0,
%\end{equation}
%with $n=1,2,3,\ldots$
%The residue of the matrix $S_{\lambda-1/2}(\omega)$ at the pole $\lambda=\lambda_n(\omega)$ is defined by [see Eq.~(\ref{Matrix_S_CAM})]
%\begin{equation}\label{residues_RP}
%r_n(\omega)=e^{i\pi [\lambda_n(\omega)+1/2]} \left[ \frac{A_{\lambda -1/2}^{(+)}(\omega)}{\frac{d}{d \lambda}A_{\lambda -1/2}^{(-)}(\omega)}\right]_{\lambda=\lambda_n(\omega)}.
%\end{equation}
%These residues play a central role in the complex angular momentum paradigm.
%

@@

\begin{figure*}[htp]
\centering
 \includegraphics[scale=0.50]{Scattering_Cross_Section_R_2-dot-26_2Mw_3}
  \vspace*{-0.3cm}
\caption{\label{S_0_R_2-dot-26_2Mw_3_Exact_vs_CAM} The scalar cross section of an ultra-compact body for $2M\omega=3$ and $R=2.26M$ and its Regge pole approximation.  The plots show the effect of including successively more Regge poles (plots 1--3). In the final plot, the sum over Regge poles gives a cross section which agrees well with the (regularized) partial-wave sum for intermediate and large values of the scattering angle.}
\end{figure*}

\begin{figure*}[htp]
\centering
 \includegraphics[scale=0.50]{Scattering_Cross_Section_R_2-dot-26_2Mw_6}
  \vspace*{-0.3cm}
\caption{\label{S_0_R_2-dot-26_2Mw_6_Exact_vs_CAM} The scalar cross section of an ultra-compact body for $2M\omega=6$ and $R=2.26M$ and its Regge pole approximation. The plots show the effect of including successively more Regge poles (plots 1--3). In the final plot, the sum over Regge poles gives a cross section which agrees well with the (regularized) partial-wave sum for intermediate and large values of the scattering angle.}
\end{figure*}

@@

Figures \ref{Rainbow_Cross_Section_R_6_2Mw_16} and \ref{Diff_Contribution_Cross_Section_R_2-dot-26_2Mw_6} show the relative magnitudes of the contributions from the different branches of Regge poles. For $R/M = 6$ (Fig.~\ref{Rainbow_Cross_Section_R_6_2Mw_16}) the surface-wave and broad-resonance amplitudes are similar in magnitude, and similar in magnitude to their sum. There is a difference in phase between these amplitudes which generates the peaks and troughs of the rainbow-scattering pattern.

@@

In this work, we have: (1) calculated the full spectrum of Regge poles for a compact body spacetime for the first time; (2) applied CAM theory to obtain an expression for the scattering amplitude as a residue series accompanied by a background integral, Eqs.~(\ref{CAM_Scalar_Scattering_amp_tot})--(\ref{CAM_Scalar_Scattering_amp_decomp_RP}); and (3) utilised the CAM formulae to calculated scattering cross sections numerically, demonstrating precise agreement with the numerical results of the partial-wave expansion, first computed in Ref.~\cite{Dolan:2017rtj}.

@@

By applying a WKB method at lowest order, we have obtained a better understanding of the spectrum of the broad resonances, finding that the magnitude of the imaginary part is linked to the magnitude of the discontinuity in the effective potential at the surface of the body (see Eq.~(\ref{eq:alpha}), Eq.~(\ref{lambda_Approx}) and Fig.~\ref{fig:Veff}). Extending the WKB method to higher orders would improve the accuracy of this approximation, and further work is need to obtain asymptotic approximations for the surface-wave branch, and the narrow-resonance branch.

We can be confident that there are no additional branches of Regge poles in the first quadrant, for two reasons. First, we have scanned across the complex-$\lambda$ domains shown in Figs.~\ref{RP_R6_approx_2Mw_3_6_s_1} and \ref{RP_R226_approx_2Mw_3_6_s_1}, as detailed in Sec.~\ref{subsec:method}. Second, in our calculation of $d\sigma/d\Omega$, we find a precise numerical agreement between the CAM calculation and the partial-wave calculation; this would not be the case if poles in this domain had been missed.

The CAM toolkit was applied here to gain a complementary practical method for calculating scattering amplitudes. However, in its current form, the method is not particularly efficient, for two reasons; first, it requires the computation of a background integral in addition to residue sum; and second, the poles on the broad-resonance branch have an approximately uniform spacing, $\Delta \lambda \sim 2$, and thus a large number of poles must be calculated. In the black-hole case, by contrast, there is only a single branch of surface-wave-type modes (see Fig.~\ref{RP_R6_approx_2Mw_3_6_s_1}), and in the high-frequency limit, a single pole captures both the orbiting and glory phenomena \cite{Folacci:2019cmc, Folacci:2019vtt}. In the case of Mie scattering, a powerful extension of the CAM approach is to define the so-called Regge-Debye poles \cite{Nussenzveig:2006}, thereby eliminating the surface-wave branch entirely. This is a possible direction for future investigation.

%In this work, we have  $R=2.26M$, which is above the Buchdahl bound but nevertheless below ... blah blah, work in Tom's text.
%Remarkably, in the 1970s, experimental data on rainbows in nucleon-nucleon scattering experiments was used to constrain models of the nuclear potential.
%Finally a paragraph on the connections to Mie scattering by a transparent sphere.

\acknowledgments
M.O.E.H. wishes to thank Antoine Folacci for various discussions concerning this work and thanks S.R.D.~for his kind
invitation to the University of Sheffield. T.S.~acknowledges financial support from EPSRC. S.R.D.~acknowledges financial support from the European Union's Horizon 2020 research and innovation programme under the H2020-MSCA-RISE-2017 Grant No.~FunFiCO-777740, and from the Science and Technology Facilities Council (STFC) under Grant No.~ST/P000800/1.
